---
title: "Practicum 10: Model Evaluation"
author: "Data Science for Biomedical Informatics (BMIN503/EPID600)"
output: 
  html_document:
    toc: true
    toc_float: 
        collapsed: true
        smooth_scroll: true
    depth: 3 
    theme: paper 
    highlight: tango
---

```{r global options, include = FALSE}
knitr::opts_chunk$set(warning = FALSE, message = FALSE)
```   
***
New packages to install for this practicum:
```{r, eval=FALSE}
install.packages("GGally")
install.packages("pROC")
install.packages("PRROC")
```
***

We will create logistic regression models to predict titanic survivorship using a dataset about [Titanic passengers](http://biostat.mc.vanderbilt.edu/wiki/pub/Main/DataSets/titanic.html) obtained from a repository hosted by [Vanderbilt Biostats](http://biostat.mc.vanderbilt.edu/wiki/Main/DataSets). The data dictionary to understand file variables is [here](https://github.com/HimesGroup/BMIN503/blob/master/DataFiles/titanic3.md). First, let's get the dataset and clean it. To look quickly at the relationships among variables graphically, we will use a command from the `GGally` package, so install that package if necessary.

```{r eval = TRUE, message = FALSE}
library(tidyverse)
library(GGally)
titanic <- read.csv(url("https://raw.githubusercontent.com/HimesGroup/BMIN503/master/DataFiles/titanic3.csv"), header=TRUE)
titanic <- titanic %>%
    select(survived, pclass, sex, age, sibsp, parch) %>%
    filter(complete.cases(.)) %>%
    mutate(survived = factor(survived, levels = c(0, 1), labels = c("died", "lived"))) %>%
    mutate(pclass = factor(pclass, levels = c(1, 2, 3), labels = c("first", "second", "third"))) 
str(titanic)
ggpairs(titanic)
ggpairs(titanic, mapping = aes(col = survived), columns = c("survived", "pclass", "sex", "age"))
```


### Logistic regression 
We create the logistic regression model and then obtain the probabilities of surviving according to the training data.
```{r eval=TRUE}
titanic.glm <- glm(survived ~ ., data = titanic, family = binomial(logit))
summary(titanic.glm)
glm.pred <- predict(titanic.glm, titanic, type = "response")
```

### Cross-Validation
When independent data is not available to test a model, we can use cross-validation, which consists of using portions of the data at a time to train and test a model. Prediction error is better estimated using cross-validation than on the same data used to train a model, where there is a risk of having an overestimate due to overfitting.

```{r eval = TRUE, message = FALSE}
#K-Fold Cross Validation
N = nrow(titanic)
K = 10
set.seed(1234)
s = sample(1:K, size = N, replace = T)
pred.outputs.glm <- vector(mode = "numeric", length = N)
obs.outputs <- vector(mode = "numeric", length = N)
offset <- 0
for(i in 1:K){
	train <- filter(titanic, s != i)
	test <- filter(titanic, s == i)
  obs.outputs[1:length(s[s == i]) + offset] <- test$survived
    
  #GLM train/test
	glm <- glm(survived ~ ., data = train, family = binomial(logit))
  glm.pred.curr <- predict(glm, test, type = "response")
  pred.outputs.glm[1:length(s[s == i]) + offset] <- glm.pred.curr

	offset <- offset + length(s[s == i])
}
```


### ROC Curves
ROC Curves provide an intuitive display of how well a predictive test performs over all possible thresholds that can be used to divide two outcomes. One R package that can create ROC curves and compute associated areas under the ROC curves (AUCs) and confidence intervals is `pROC`.
```{r eval = TRUE, message = FALSE}
library(pROC)
roc(titanic$survived, glm.pred, ci = TRUE)
plot.roc(titanic$survived, glm.pred, ci = TRUE)
plot.roc(obs.outputs, pred.outputs.glm, col = "red", add = TRUE)
glm.roc <- plot.roc(titanic$survived, glm.pred, col = "red", lwd = 3, grid = TRUE)
plot(ci.thresholds(glm.roc), col = "grey")
legend("bottomright", legend = c("Training", "Cross-Validation"), col = c("black", "red"), lwd = 2)
```

### PR Curves
Precision-recall curves are an alternative to ROC curves that perform better with class imbalance. In the worst case, their plot is a horizontal line equal to the number of true positive tests over all training points. In the best case, they form a square (horitonal line to (1, 1). The area under the curve (AUC) is a summary measure of the curve.
```{r eval = TRUE, message = FALSE}
library(PRROC)
roc <- roc.curve(glm.pred[titanic$survived == "lived"], glm.pred[titanic$survived == "died"], curve=TRUE)
plot(roc)
roc.cv <- roc.curve(pred.outputs.glm[obs.outputs == "2"], pred.outputs.glm[obs.outputs == "1"], curve=TRUE)
plot(roc.cv)
pr <- pr.curve(glm.pred[titanic$survived == "lived"], glm.pred[titanic$survived == "died"], curve=TRUE)
plot(pr)
pr.cv <- pr.curve(pred.outputs.glm[obs.outputs == "2"], pred.outputs.glm[obs.outputs == "1"], curve=TRUE)
plot(pr.cv)
```

