---
title: 'Practicum 15: Spatial Data and Mapping'
author: 'Data Science for Biomedical Informatics (BMIN503/EPID600)'
output: 
  html_document:
    toc: true
    toc_float: 
        collapsed: false
        smooth_scroll: true
    depth: 4 
    theme: paper 
    highlight: tango
---

```{r global options, include=FALSE}
library(ggmap)
knitr::opts_chunk$set(warning=FALSE, message=FALSE)
```   
***
New packages to install for this practicum:
```{r, eval=FALSE}
install.packages("sp")
install.packages("rgdal")
install.packages("sf")
install.packages("tidycensus")
install.packages("ggsn")
install.packages("mapview")
install.packages("leaflet")
```
Today, we'll be querying Census data using _tidycensus_, which will require that you get a Census API key. You can get one as follows:

1. Go to https://www.census.gov/developers/

2. Click on the "Request a KEY" box on the left side of the page.

3. Fill out the pop-up window form.

4. You will receive an email with your key code in the message.

Once you have received your API key, it can be set using the `census_api_key` command. Note this only needs to be done once, as we will show below.

***

## Spatial data overview

First, let's understand how spatial data is handled in R. Historically speaking, `sp` classes and methods have been the backbone of spatial tools in R, but the `sf` framework is rapidly being adopted due to its user-friendliness and compatibility with the tidyverse suite of packages (including _dplyr_ and _ggplot2_).

Let's examine how R handles different types of GIS data, beginning with vector data (i.e. points, lines and polygons) and ending with a brief raster example. We will initially compare the `sp` and `sf` frameworks in our line data example, but the majority of the practicum will take place within the `sf` framework.

### Line data 

We will begin with the [Philadelphia Bike Network](https://www.opendataphilly.org/dataset/bike-network) as an example of line data. This dataset includes all streets in Philadelphia with bicycle access. We will initially handle this data as an `sp` object, followed by `sf`, in order to compare these two frameworks.

First let's download the data in shapefile format. 

```{r eval=TRUE, message=FALSE}

# Download the Philadelphia Bike Network shapefile from OpenDataPhilly
dir.create("bike_network")  # create directory (folder) to store our shapefile
setwd("bike_network")       # navigate to the bike_network directory
download.file("http://data.phl.opendata.arcgis.com/datasets/b5f660b9f0f44ced915995b6d49f6385_0.zip", "bike_network.zip")  # download zipped shapefile from OpenDataPhilly
unzip("bike_network.zip")  # unzip the file
if (file.exists("bike_network.zip")) file.remove("bike_network.zip")
setwd("..")                # navigate back to parent folder (good practice so you can easily find output later)

```

Next, let's read the shapefile into R as an `sp` object so that we can get a better understanding of the `sp` framework. 
```{r}
library(sp)
library(rgdal)

bn.sp <- readOGR(dsn="bike_network", layer="Bike_Network")  # read shapefile as an sp object
class(bn.sp)             # bn.sp is a SpatialLinesDataFrame
plot(bn.sp)              # sp objects can be mapped directly using the base plot command
str(bn.sp, max.level=2)  # max.level makes read-out more manageable
head(bn.sp@data)         # data attributes stored in the data slot in data.frame form
str(bn.sp@lines, max.level = 2,  # geometry attributes stored in lines slot
    list.len = 10)       # list.len sets number of list elements to display
bn.sp@lines[[1]]         # double brackets access the first line of the lines slot
bn.sp@lines[[2]]         # geometry attributes of the second line
bn.sp@data[1:2,]         # these lines correspond to segments of Spring Garden Street
bn.sp@bbox               # boundaries stored in bbox slot (bbox = bounding box)


```

Now, let's compare the above to how spatial data is handled in the `sf` framework. The advantage of working within the `sf` framework is that `sf` objects are a type of `data.frame` and can thus be easily manipulated with standard subsetting and data wrangling commands.

```{r}
library(sf)

bn.sf <- st_read("bike_network/Bike_Network.shp")  # read shapefile as an sf object
class(bn.sf)  # bn.sf is an sf object, which is a subclass of data.frame

# sf objects can be handled like data frames using familiar commands
str(bn.sf)
head(bn.sf)  
dim(bn.sf)
bn.sf[1,]  
head(bn.sf$STREETNAME)  
head(bn.sf[,3])

# Spatial attributes of sf objects can be accessed with the st_geometry command
bn.geo <- st_geometry(bn.sf)
plot(bn.geo)
bn.geo[[1]]  # line segment 1 
bn.geo[[2]]  # line segment 2
bn.sf[1:2,]

```

Due to it's compatibility with the tidyverse and the opinion of many experts that `sf` will ultimately supersede `sp`, we will only work within the `sf` framework for the remainder of this practicum.

### Polygon data
We will look at Philadelphia census tracts as an example of polygon data. Census shapefiles are free to download on the [Census website](https://www.census.gov/geo/maps-data/data/tiger-line.html), but we will load a version that has already been converted to an sf object. 

```{r}

# Load philly tracts data
philly.tracts <- readRDS(gzcon(url("https://raw.githubusercontent.com/HimesGroup/BMIN503/master/DataFiles/philadelphia_tracts_2010.rds")))

# Note philly.tracts is an sf ("simple feature") object of type "MULTIPOLYGON"
philly.tracts  
class(philly.tracts)  

# Once again, we can extract the geometry of philly.tracts with the st_geometry function
pt.geo <- st_geometry(philly.tracts)
pt.geo
pt.geo[[1]]        # perimeter coordinates for the first census tract of the sf
philly.tracts[1,]  # i.e. Census Tract 94

pt.geo[[2]]        # perimeter coordinates for the second census tract of the sf
philly.tracts[2,]  # i.e. Census Tract 95

# Plot the geometry of philly.tracts with the base plot function
plot(pt.geo)

# The base plot function has some aesthetic options we can use to tweak our plots
plot(pt.geo, col = "lemonchiffon2")
plot(pt.geo, lwd = 2, border = "red")

```


### Point data
As an example of point data, we will work with crime incidents that occurred in Philadelphia in September 2018. The full publicly available crime incidents database for Philadelphia is maintained by the Philadelphia Police Department and is available on the [OpenDataPhilly](https://www.opendataphilly.org/dataset/crime-incidents) website.

```{r}
library(tidyverse)

crime <- readRDS(gzcon(url("https://raw.githubusercontent.com/HimesGroup/BMIN503/master/DataFiles/crime_incidents.rds")))

# The crime data is an sf object of type POINT and includes information on the date, time and offense type for each incident
crime
str(crime)  

# Factor or character variables in a standard date format can be converted to class "Date" with as.Date
summary(crime$dispatch_date)
crime$dispatch_date <- as.Date(crime$dispatch_date)  
summary(crime$dispatch_date)

# There are certain perks that come with the "Date" class
crime$day_of_week <- weekdays(crime$dispatch_date)
head(crime)
hist(crime$dispatch_date, breaks = "days")

# Let's take a look at offense types and use dplyr to filter by offense_type...
table(crime$offense_type)
homicide <- filter(crime, offense_type == "Homicide - Criminal")
fraud <- filter(crime, offense_type == "Fraud")

# Note subsets of an sf object are also sf objects
class(homicide)
class(fraud)

# Plotting homicide and fraud incidents with the base plot function
plot(st_geometry(homicide))

# Points by themselves are not very easy to understand. Let's layer them on top of the tract polygons with add = TRUE
plot(pt.geo)
plot(st_geometry(fraud), col = "blue", alpha = 0.1, add = TRUE)
plot(st_geometry(homicide), col = "red", add = TRUE)
legend("bottomright", legend = c("Fraud", "Homicide"), title = "Offense type:", col = c("blue", "red"), pch = 1, bty = "n")


```


### Raster data
So far, we have considered point, line, and polygon data, all of which fall under the umbrella of vector data types. Rasters are a distinct GIS data type that we will consider only briefly because they cannot be handled with `sf` methods. We will look at the `volcano` dataset, which gives topographic information for Maunga Whau (a volcano located in Auckland, New Zealand) on a 10m by 10m grid. Because it is a relatively small raster, we can handle `volcano` using base functions. Larger rasters should be handled using the _raster_ package.

```{r}

library(datasets)

# The volcano dataset is a 87x61 matrix
class(volcano)
str(volcano)
filled.contour(volcano, color = terrain.colors, asp = 1)
```



## Census data (_tidycensus_)

Data from the Decennial Census and the American Community Survey (ACS) can be queried directly from R with the _tidycensus_ package. _tidycensus_ interfaces with Census APIs and returns Census data in tidyverse-ready data frames. You should have a Census API key, having followed the instructions at the top of this practicum.

```{r}
library(tidyverse)
library(tidycensus)

census_api_key("ee0e6e63a6ec3ae184a66061270249f71f84cf35", install=TRUE)

```

### Searching for Census variables

Getting select variables from the Census or ACS requires knowing the associated variable IDs among the thousands available across the different Census files. To rapily search for variables, use the `load_variables` function. Then use CTRL/COMMAND-F to search for variables by name. 

```{r}

vars <- load_variables(dataset = "acs5",  # specify ACS 5-year estimates 
                       year = 2016)  # specify end year (i.e. this queries ACS 5-year estimates for 2011-2016)


View(vars)
```

### Accessing Census data

The `get_decennial` function grants access to the 1990, 2000, and 2010 decennial US Census APIs and the `get_acs` function grants access to the 5-year ACS 

```{r}

poverty <- get_acs(geography = "tract",         # query data at the tract level 
                   year = 2016,                 # end year (these will give us ACS 5-year estimates for 2011-2016)
                   variables = c("B17010_002",  # number of families falling below the poverty threshold
                                 "B17010_001"), # total number of families for which poverty was determined
                   state = "PA", county = "Philadelphia")

head(poverty)

percent.poverty <- poverty %>%
  group_by(GEOID) %>%  # return only one row per GEOID
  mutate(pov_fam = estimate[variable == "B17010_002"],  # extract and rename number of families in poverty 
         tot_fam = estimate[variable == "B17010_001"],  # extract and rename total number of families
         percent_poverty = (pov_fam/tot_fam)*100) %>%   # multiply by 100 to get rate as a percentage
  select(GEOID, NAME, pov_fam, tot_fam, percent_poverty) 

head(percent.poverty)
```

We have tract-level poverty rates for Philadelphia, which is great! However, this data is hard to comprehend in tabular format. Next, let's visualize these poverty rates on a map. To do this, we'll have to merge our table of poverty rates with our census tract data.

### Joining data to sf objects

Because `sf` objects are a type of `data.frame`, they can be readily merged with tabular data (e.g. from ACS)- so long as the latter contains an index variable on which to perform the merge. Let's merge our poverty rates data frame with our census tract sf object using a _dplyr_ join function.

```{r}

# Before performing our join, we need to determine which variable to join on. 
head(philly.tracts$GEOID10)
head(percent.poverty$GEOID)

# Both philly.tracts and percent.poverty have a Census GEOID variable with matching foramt, but they have slightly different names. Thus, we'll need to rename one of them so they match.
philly.tracts <- mutate(philly.tracts, GEOID = as.character(GEOID10))
  
# Now we can perform our join
philly.tracts2 <- inner_join(philly.tracts, percent.poverty, by = "GEOID")
head(philly.tracts2)  # success!

```

In the next section, we wil use _ggplot2_ to visualize tract-level poverty rates in Philadelphia and see if these might be related to homicide locations.

## Static maps

### Mapping with _ggplot2_

`sf` objecs can be easily plotted with _ggplot2_.
```{r}

# Note we can plot sf objects directly (no need to convert to sfc using st_geometry)
ggplot() +
  geom_sf(data = philly.tracts)

# We can layer sf objects together
ggplot() +
  geom_sf(data = philly.tracts) +
  geom_sf(data = bn.sf, color = "blue") +
  geom_sf(data = homicide, color = "red")

```

We can make a choropleth map of poverty rates in Philadelphia census tracts with the `fill` aesthetic.

```{r}
ggplot() +
  geom_sf(data = philly.tracts2, aes(fill = percent_poverty))

```

### Customization in _ggplot2_

Next, let's add some polish to our map with _ggplot2_'s `theme()` commands and custom color palettes. 

```{r}
library(RColorBrewer)
library(viridis)

myPalette <- colorRampPalette(brewer.pal(9, "BuPu"))  # RColorBrewer palette

# To save space, we can save theme commands as a function that we can "add" to calls to ggplot
my_theme <- function() {
  theme_minimal() +                                  # shorthand for white background color
  theme(axis.line = element_blank(),                 # further customization of theme components
        axis.text = element_blank(),                 # remove x and y axis text and labels
        axis.title = element_blank(),
        panel.grid = element_line(color = "white"),  # make grid lines invisible
        legend.key.size = unit(0.8, "cm"),           # increase size of legend
        legend.text = element_text(size = 16),       # increase legend text size
        legend.title = element_text(size = 16))      # increase legend title size
}

ggplot() +
  geom_sf(data = philly.tracts2, aes(fill = percent_poverty)) +
  geom_sf(data = homicide, color = "gold", size = 3) +
  my_theme() +                                           # add our custom theme
  ggtitle("Tract-level poverty rates in Philadelphia") + # add plot title
  scale_fill_gradientn(name = "Poverty \nrate (%)",      # change legend title
                    colours = myPalette(100))            # RColorBrewer palette OR
                    #colours = viridis(100))             # viridis palette
  
```

### Scalebars and north arrows

Publication-ready maps should have scalebars and north arrows, which can be added to maps made with _ggplot2_ with commands in the _ggsn_ library.

```{r}
library(ggsn)

ggplot() +
  geom_sf(data = philly.tracts2, aes(fill = percent_poverty)) +
  my_theme() +    
  ggtitle("Tract-level poverty rates in Philadelphia") + 
  scale_fill_gradientn(name = "Poverty \nrate (%)",  
                    colours = myPalette(100)) +
  north(x.min = -75.28031, y.min = 39.86747,    # add north bar with north()
        x.max = -74.95575, y.max = 40.13793,    # set map boundaries with x.min, y.min, etc..      
        symbol=12,                              # select north arrow symbol
        anchor = c(x = -75, y = 39.93)) +       # set north bar location
  scalebar(x.min = -75.28031, y.min = 39.86747, # add scalebar with scalebar()
           x.max = -74.95575, y.max = 40.13793, # set map boundaries
           dist = 5, dist_unit = 'km',          # set scalebar segment length = 5km
           transform = TRUE,                    # TRUE for decimal degree coordinates
           model = "WGS84")                     # specify CRS
                       

```

### _ggmap_

Currently, _ggmap_ does not support `sf` objects. To make a map that incorporates map tiles from _ggmap_, one needs to start with an `sp` object and then convert to a `data.frame` object. As an example, I have included code that maps tract-level poverty rates in Philadelphia over a minimal basemap. _ggmap_ interfaces with the GoogleMaps API and requires users to register an API key with GoogleMaps. Instructions for obtaining a GoogleMaps API key can be found [here](https://developers.google.com/maps/documentation/embed/get-api-key).

```{r eval = FALSE}
library(ggmap)
register_google("YOUR_API_KEY_HERE")

# Download Philadelphia census tracts as an sp object
philly.tracts.sp <- readRDS(gzcon(url("https://raw.githubusercontent.com/HimesGroup/BMIN503/master/DataFiles/philadelphia_tracts_2010_sp.rds")))

#  Convert sp object to data frame, so it can be plotted with ggplot2
pt.df <- fortify(philly.tracts.sp, region = "GEOID10")
pt.df <- rename(pt.df, GEOID = id)

# Merge pt.df with poverty rates data frame 
head(pt.df)
pt.df <- merge(pt.df, percent.poverty, by = "GEOID")
head(pt.df)

# Retrieve map tile with qmap function
philly.toner <- qmap("19133", maptype = "toner-lite", zoom = 11)

philly.toner + 
  geom_polygon(data = pt.df, aes(x = long, y = lat, group = group, # must specify long/lat and group variables
                                 fill = percent_poverty),
               alpha = 0.7) +                                      # alpha sets transparency of polygons over basemap
  my_theme() +
  scale_fill_gradientn(name = "Poverty \nrate (%)",
                       colours = myPalette(100))


```




## Interactive maps

Interactive maps invite viewers to engage with geospatial data on a deeper level by allowing them to pan around and zoom into any part of a geographic dataset overlaid on a basemap to show context. We can also incorporate popup messages into interactive maps to give viewers additional information about each map feature as they click on it.

### Interactive maps with _mapview_

The _mapview_ library uses the _leaflet_ javascript library to quickly and easily create interactive maps from `sf` objects. Popup messages with data attributes associated with each map feature are automatically generated using the `mapview` command. The disadvantage of using _mapview_ is that it has few customizable options. Thus, you may want to use `mapview` to do a quick exploration of geospatial data before creating a more custom map using the R _leaflet_ library. 

```{r}
library(mapview)

mapview(homicide)
###Can't really customize this map
```

### Interactive maps with _leaflet_

The _leaflet_ library allows users to interface with the javascript library of the same name to create custom interactive maps.

```{r}
library(leaflet)

# Select a color palette with which to run the palette function
pal_fun <- colorNumeric("BuPu", NULL)       # Blue-Purple from RColorBrewer
pal_fun2 <- colorNumeric("YlOrRd", NULL)    # Yellow-Orange-Red from RColorBrewer
pal_fun3 <- colorNumeric("viridis", NULL)   # viridis from viridis
pal_fun4 <- colorNumeric("inferno", NULL)   # inferno from viridis
pal_fun5 <- colorNumeric("inferno", NULL, reverse=TRUE)  # reverses the color ramp

# Pop-up message
pu_message <- paste0(philly.tracts2$NAMELSAD10,  # paste0 to append tract name with other relevant text 
                     "<br>Poverty rate: ",       # <br> forces new line
                     # use round function to round continuous poverty rate to one decimal point
                     round(philly.tracts2$percent_poverty, 1), "%")

# Basic leaflet map
leaflet(philly.tracts2) %>%
  addPolygons(fillColor = ~pal_fun(percent_poverty), # addPolygons requires variable and a pass to a palette function  
              popup = pu_message) %>%                # add a popup message
  addTiles()                                         # default basemap

# Adding more customization 
leaflet(philly.tracts2) %>%
  addPolygons(stroke = FALSE,                        # remove polygon borders
              fillColor = ~pal_fun(percent_poverty),
              fillOpacity = 0.5, smoothFactor = 0.5, # increase opacity and resolution
              popup = pu_message) %>%
  addProviderTiles(providers$CartoDB.Positron) %>%   # add third party provider tile
  #addProviderTiles(providers$Stamen.Toner) %>%
  #addProviderTiles(providers$Esri.NatGeoWorldMap)
  addLegend("bottomright",                           # location of legend
            pal=pal_fun,                             # palette function
            values=~percent_poverty,                 # variable to be passed to palette function
            title = 'Poverty rate',                  # legend title
            opacity = 1) %>%                         # legend opacity (1 = completely opaque)
  addScaleBar()

```




