---
title: 'Practicum 3: R Programming Basics'
author: 'Data Science for Biomedical Informatics (BMIN503/EPID600)'
output: 
  html_document:
    toc: true
    toc_float: 
        collapsed: true
        smooth_scroll: true
    depth: 3 
    theme: paper 
    highlight: tango
---

***
In this practicum, we will become familiar with basic R data types and commands. Learning these is necessary to perform more advanced analyses. If you are familiar with programming, then this will be a straightforward exercise, if not, then hang in there and learn as much as you can. The time you invest in learning R will pay off later. From now on, it will be convenient to download the `Rmd` versions of these practicum files so that you can run code directly, rather than have to type in or copy/paste from the `html` documents.


### RStudio
As discussed in class, RStudio is a free R user interface that makes development of code and integration of R files with reproducible research tools very straightforward. Explore the RStudio windows. In the R Console, you can type and execute R commands. For example, try these commands in the Console (if looking at the html doc, type in the grey portions and your output should be what appears in the white boxes):
```{r, eval=TRUE}
1+2
c(1:5)
cat("hello", "there")
x <- rnorm(1000)
hist(x)
```

The Terminal window is the one we've used to execute git commands. We will discuss R commands in more detail later. Above the R Console, you can edit files with R code or create text files as above. To the right, there are windows that show variables loaded, history of commands, connections (e.g., via git), file path navigation, R plot output, package installation, and R documentation.


### R Settings
Prior to working in R, there are some Console commands that you should become familiar with. The current _working directory_, where your data is read and saved to can be checked by typing the follwing in the Console:
```{r, eval=FALSE}
getwd()
```
To see what files are in this current directory:
```{r, eval=FALSE}
dir()
```
To see what functions are loaded:
```{r, eval=FALSE}
ls()
```

To change directories, you can use `setwd()` within the command line or in an R script. With RStudio, we can use GUI tools to do all of the above. For example, you can change directories by browsing to one: go to Session -> Set Working Directory -> Choose Directory, then select a new directory. Alternatively, you can use the "Files" panel.


### R Packages
As mentioned during the lecture, many of the R functions you will use are in packages, which have to be installed and loaded prior to use. From the R console, install these packages from CRAN:
```{r, eval=FALSE}
install.packages("knitr")
install.packages("rmarkdown")
install.packages("tidyverse")
```

In RStudio, packages can be installed interactively with the "Install Packages" tool in the "Packages" tab. Once packages are installed, it is necessary to tell R about them each time you start an R session or include them within a script where they are used with the `library()` command:
```{r, eval=FALSE}
library(tidyverse)
```
To see loaded packages, use the `search()` funcion.

### Help in R
To get help about specific R functions, type `?` before the command. In RStudio, the effect of this will be to open the `Help` window with documentation for that specific command.
```{r, eval=FALSE}
?read.table
```
More broadly, when you don't know the name of the command you are looking for, you can use the following command to do a text search through the R documentation:
```{r, eval=FALSE}
help.search("mean")
```

Recall that web searches often locate answers to problems encountered or can be used to look up functions. Use `sessionInfo()` when asking a question about R either by email or to a forum. This helps others understand the environment where a problem arises.


### R Scripts
Any text editor can be used to edit R files, including the one that is available via the R console. We will use RStudio to edit text files with R commands. _R script_ files (.R ending) consist of R functions, comments, and variables. To create a new R script file, use the RStudio new document icon in the upper-left-hand corner and select "R Script". In the new document, type the following:
```{r, eval=FALSE}
#Test R script
x<-1
y<-2
x+y
```
Now save the file as `test.R`. To execute the code, click on "Run" above the text editor window. Note that the single line where the cursor is located is run by default. You can select multiple lines of code and then click "Run" to execute more than one line. To run the entire document, you can select "Source" instead. Alternatively, you can load an R script into your R environment at the command line by typing: 
```{r, eval=FALSE}
source("test.R")
```
Verify that the contents of `test.R` loaded properly by typing `x` and `y` at the command line. Now modify their values in the document and repeat the steps to see their values changed.


### R Markdown
For class, we will mostly use R Markdown files (.Rmd ending) since these are easily converted into visually appealing output documents that include text, graphics, and R code. Such documents are a great way to keep a "lab notebook" for data science projects. To create a new R Markdown file, use the RStudio new document icon in the upper-left-hand corner and select "R Markdown". Give it a title and author of your choice (or leave the default options) and click "OK". The new document has some sample text and commands that illustrate some of the main components of an R Markdown file. To save the R Markdown file, use File -> Save As, and then choose a `test.Rmd` and click "Save". Click on the "Knit HTML" icon at the top of the file to generate an html document based on the template and compare the two. If you'd like, repeat this with PDF and Word options to see other output formats. For these latter two to work, you have to have additional software installed on your computer.

You can download an R Markdown [cheat sheet](https://www.rstudio.com/wp-content/uploads/2016/03/rmarkdown-cheatsheet-2.0.pdf), find one under Help->Cheatsheets menu of RStudio, and read more about it here: http://rmarkdown.rstudio.com. To see all of the settings possible with html documents, you can read this [guide](http://rmarkdown.rstudio.com/html_document_format.html).


### Basic R Commands
To assign values to variables, use `<-`, the assignment operator. You will also see `=` used in several cases, especially to assign entries within a braced list of expressions or function to avoid global assignment of variables. At the prompt, `<-` and `=` will behave equally.
```{r eval=FALSE}
x <- 1
y <- c(1,4,8)
print(x) 
print(y)
x 
y
```

Notice the output of `x` and `print(x)` are the same. Typing `x` is called _auto-printing_, while `print(x)` is _explicit printing_. Sometimes explicit printing is needed if the default behavior of a call is not to print, for example, within functions or to output graphics - more on this later. You can use `#` as a prefix to insert comments into code.
```{r eval=TRUE}
#List of colors:
my.colors <- c("red", "yellow", "blue")
print(my.colors)
```

Incomplete entries on the R prompt continue onto another line.

To get a list of integers use `:`. To create lists of numbers with more flexibility use `seq`. To create multiple instances of a pattern, use `rep`.
```{r eval=TRUE}
x <- 1:50
x 
x2 <- seq(1, 10, 2)
x2
x3 <- rep(1, 20)
x3
```

### R Data Types
There are five basic objects in R:

- Character (i.e. text)
- Numeric (i.e. real numbers)
- Integer
- Complex
- Logical (TRUE/FALSE)

To find out the type of object a variable x is, use `class(x)`. Each object can have atrtributes, including:

- Names
- Dimension
- Class
- Length

To find out what other attributes a particular object has, use `attributes(x)`
```{r eval=TRUE}
class(x)
class(colors)
class(Inf) #Infinity
class(1)
class(1L) #Use the letter L after a number to specify that it is an integer
```
Missing values in R are denoted as `NA`, while undefined numeric values are denoted `NaN`. To find out if values are missing, use `is.na()`. Similarly, `is.nan()` can be used to test for `NaN`. `NA` can take on the class of the members of its data structure.
```{r, eval=TRUE}
class(NaN) #Undefined numeric value "not a number"
0/0
x <- 1
is.na(x)
!is.na(x) #Notice negation of is.na() by preceding with !
y <- NA
is.na(y)
!is.na(y)
```


### Data Structures
#### Vectors
Vectors, can be created with the `c()` function or the `vector()` function. Its contents can be of one type or a mixture. In the case when there is a mixture of classes, all values will be coerced into a common class. To force a change to a different class, use functions such `as.numeric()`, `as.character()`.
```{r, eval=TRUE}
x <- c(1:5)
y <- c("a", 1, TRUE)
z <- c(x, y)
print(x)
print(y)
print(z)
print(c(class(x), class(y), class(z)))
as.character(x) #A coercion that may be helpful
as.numeric(y) #A coercion that may not make sense
```


#### Lists
Lists are similar to vectors, but they can contain elements of different classes.
```{r, eval=TRUE}
x <- list(1, "a", c(1:5))
print(x)
print(class(x))
print(class( x[[1]] ))
print(class( x[[2]] ))
print(class( x[[3]] ))
```


#### Factors
Factors are similar to vectors, but they represent categorical data. Often used for statistical modeling since variables can keep an intuitive label, rather than have to be converted to numbers first. They can be ordered or unordered.
```{r, eval=TRUE}
x <- c("male", "male", "female", "female", "male")
x
table(x)
y <- factor(x)
#Notice the "Levels:" after the output
y 
table(y)
```
Although the factor looks like a text list, the underlying structure in R is numerical, which is necessary for use of statistical models. The underlying structure can be viewed with `unclass()`.
```{r, eval=TRUE}
unclass(y)
```
The order of factors can be set using the `levels` argument. The first argument is used as _baseline_ level when making statistical models. We will come back to this in a future practicum.
```{r, eval=TRUE}
y2 <- factor(c("male", "male", "female", "female", "male"), levels=c("male", "female"))
y2
```


#### Matrices
Marices are vectors with a dimension attribute that contains the number of rows and columns. Elements must be of the same class.
```{r, eval=TRUE}
x <- matrix(nrow=3, ncol=4)
x
dim(x)
```
A common way to create matrices or add rows/columns is the use of `cbind()` or `rbind()`.
```{r, eval=TRUE}
x <- c(1:4)
y <- c(5:8)
m <- cbind(x,y)
m2 <- rbind(x,y)
x
y
m
m2
```


#### Data Frames
Data frames are like lists, in that they can be composed of entries of multiple classes, but they are 2-dimensional. This data structure is used very often for data analysis in R, and it is a format we will most often during the class. Each column of a data frame must have elements of a single class, and each column must be of the same length.
```{r, eval=TRUE}
x <- data.frame(c(1:5), c("a","b","c","d","e"))
x
names(x) <- c("Numbers", "Letters")
x
dim(x)
```
Names are most commonly used with data frames, but these can also be helpful when used with other data types (i.e. vectors, lists, matrices).

#### Diagnostics
If you are working with an object and need to find out more about it, the `summary()` and `str()` functions come in handy. The `str()` command will show a brief display of an object's internal structure or give informatin about a function. The `summary()` command gives output that attempts to summarize variables.
```{r, eval=TRUE}
str(x)
summary(x)
str(y2)
summary(y2)
str(data.frame)
str(read.table)
```

***

