---
title: 'Practicum 4: R Programming Basics II'
author: 'Data Science for Biomedical Informatics (BMIN503/EPID600)'
output: 
  html_document:
    toc: true
    toc_float: 
        collapsed: true
        smooth_scroll: true
    depth: 3 
    theme: paper 
    highlight: tango
---

***

In this practicum, we will continue to become familiar with R by learning about some programming basics, saving R objects, retrieving portions of data objects, and creating functions.


### Programming Basics
R can be used to write simple to advanced functions. Doing this requires learning commands that are part of the language. Most R users have to write scripts with functions to analyze data, even if not methodologically advanced. This saves time and helps with reproducibility. More advanced users may write R packages with novel methods or make current methods more efficient. Code from other languages, such as C/C++ can be called into R code to make complex code run more efficiently. For this class, we will cover some of the basic programming structures that will allow you to write scripts to analyze data. As you become familiar with these, you can continue to learn more from a textbook or online resources. 

#### If Statements
```{r, eval=TRUE}
x <- 2
if(x > 10) {
  y <- 20
  print("Your number is greater than 10")
} else {
  y <- 0
  print("This is not a number greater than 10")
}
y

x <- 200
if(x > 10) {
  y <- 20
  print("Your number is greater than 10")
} else {
  y <- 0
  print("This is not a number greater than 10")
}
y
```
_If statements_ are not very helpful when written as above. Rather, they are often part of a function where a value (or values) can be passed and a condition evaluated. For example, this could be a portion of a script where the value _x_ changes and a decision is made based on its value. This way, the _if statement_ would only have to be written once. Other possibilities for _if statements_ are to include _else if_ clauses or to only test one condition. Try these:
```{r, eval=FALSE}
if(x > 10){ y <- 20 }
y
rm(y)
x<-2

if(x > 10){ y <- 20 }
y

if(x > 10) {
  y <- 20
} else if(x < 5) {
  y <- 0
} else {
  y <- 10
}
y
```

#### For Loops
To repeat a task over a sequence of numbers (or items), we can use _for loops_. Their format in R is as follows:
```{r, eval=TRUE}
for (i in 1:5) {
  print(i)
}
```
Iteration can take place over any vector. For example,
```{r, eval=TRUE}
x <- c("red", "blue", "yellow", "green")
for(i in x){
  print(i)
}
```
A function that can come in handy to loop over an integer equal to a sequence length is `seq_along()`
```{r, eval=TRUE}
for(i in seq_along(x)){
  print(i)
}
for(i in seq_along(x)){
  print(x[i])
}
```
Loops can be nested by using tabs and curly brackets to demarcate each loop.
```{r, eval=TRUE}
for(i in 1:5){
  for(j in x){
    print(paste("At number", i, "with color", j))
  }
}
```
Sometimes _for loops_ are constructed with exceptions that allow one to break out of the loop structure or skip an iteration. In R, this can be done using `next`. Say you wanted to skip `i=3` in the loop above:
```{r, eval=TRUE}
for(i in 1:5){
  if(identical(i,3)) {
    next
  } else {
    for(j in x){
      print(paste("At number", i, "with color", j))
    }
  }
}
```
The `break` function is helpful to break out of a loop because further computation is unnecessary or to generate helpful "error messages" within scripts. Say you needed for x consist of numbers, rather than strings. You could use:
```{r, eval=TRUE}
x <- c("red", "blue")
for(i in 1:5){
  for(j in x){
    if(!identical(class(j), "numeric")){
      print(paste("Error: A non-numeric value was provided:", j))
      break
    }
    print(i+j)
  }
}
```
Notice what would happen without the `break` by trying to run this:
```{r, eval=FALSE}
x <- c("red", "blue")
for(i in 1:5){
  for(j in x){
    if(!identical(class(j), "integer")){
      print(paste("Error: A non-numeric value was provided:", j))
    }
    print(i+j)
  }
}
```

Loops are used within programs, and sometimes they are a necessity. However, in R, many loop actions can be completed with a vectorized operation. These can help make code easier to read and be more computationally efficient because operations with them can be performed in parallel.
```{r, eval=TRUE}
x <- 1:5
y <- 10:14
x/y
x+y
x > 1
m1 <- matrix(1:4, 2, 2)
m2 <- matrix(c(1, 1, 0, 2), 2, 2)
m1
m2
m1*m2 #An element-wise operation
m1 %*% m2 #A matrix operation
```
Be careful with these functions and check your code as you go. Here is something unexpected that could happen:
```{r, eval=TRUE}
c(1,2,1,2) * c(1,2,3,4) #No problem
c(1,2) * c(1,2,3,4) #How might this work?
c(1,2,1) * c(1,2,3,4) #And this?
```

R has many loop functions that work as a _for loop_ but are shorter to write, making code simpler: `lappy`, `sapply`, and `apply`. Here are some examples of these:
```{r, eval=TRUE}
x <- list(c(1:10), c(11:20))
sapply(x, sum)
sapply(x, length)
y <- list(a = rep(1, 5), b = rep(2, 5))
lapply(y, sum) #list apply
sapply(y, sum) #simplified apply
m1
apply(m1, 1, sum) #computes function along margins of an array 1=rows, 2=columns
apply(m1, 2, sum)
```
Two other loop functions are `tapply`, and `mapply`. `mapply` can take in multiple lists and use items of each in a function. `tapply` is used to apply a function over subsets of a vector.


### R Objects
It can be convenient to save R objects without having to change them into a generic format so that they can be loaded quickly into R sessions. The `dump()` and `source()` commands can be used to save and retrieve, respectively, several R objects. For example,
```{r, eval=TRUE}
x <- data.frame(V1 = c(1:5), V2 = rep("Y", 5))
y <- c(1:10)
z <- "Some text."
dump(c("x", "y", "z"), file = "test_data.R")
ls()
rm(x, y, z)
ls()
source("test_data.R")
ls()
x
y
z
```
To remove all objects that are loaded, use `rm(list = ls())`.


### Subsetting
When working with data frames and other R objects, we often want to extract portions of them. This can be done with a few approaches depending on the data type. Elements of a vector can be extracted with a single bracket:
```{r, eval=TRUE}
x <- c(1:5)
x[1]
```
Elements of a list can be extracted with double brackets or using names if these exist. Names are helpful when dealing with large lists or data frames when it is difficult to remember the order of items/columns.
```{r, eval=TRUE}
x <- list(1, "a", c(1:5))
x
x[1] #Returns same class as original (i.e. a list)
x[[1]] #Returns the contents of the first item
x[[3]]
y <- list("Item1"=c(1:4), "Item2"="Some text")
y
y$Item1 #Returns the contents of the first item
y[[1]] #Same as y$Item1
y[["Item1"]] #Same as y$Item1
y["Item1"] #Returns same class as original. Same as y[1]
```
A difference between subsetting with `$` and `[[]]` is that the double brackets can be used to compute an index, while the entry after `$` must be an actual element name. 
```{r, eval=TRUE}
get.name <- "Item1"
y[[get.name]]
y$get.name
```
Note that `[]` can be used with a range (for 1D objects) or ranges (for 2D objects) of objects, while `[[]]` is used to extract single entries. If a sequence of values is passed to a double bracket, then an attempt will be made to recurse down levels.
```{r, eval=TRUE}
y[[c(1, 1)]]
y[[1]][[1]]
y$Item1[[2]]
m <- matrix(c(1:6), 2, 3)
m[1, 1]
m[1, ]
m[, 1]
```
With matrix subsetting, dimensionality of output objects is reduced by default. This can be changed with the `drop` option:
```{r, eval=TRUE}
m[1, 1]
m[1, 1, drop=FALSE]
m[1, ]
m[1, , drop=FALSE]
```
Another helpful way to select portions of a list is the use of the `%in%` operator.
```{r, eval=TRUE}
cities <- c("Juneau","Phoenix","Little_Rock","Sacramento","Denver","Hartford","Dover","Tallahassee","Atlanta","Honolulu","Boise","Springfield","Indianapolis","Des_Moines","Topeka","Frankfort","Baton_Rouge","Augusta","Annapolis","Boston","Lansing","St._Paul","Jackson","Jefferson_City","Helena","Lincoln","Carson_City","Concord","Trenton","Santa_Fe","Albany","Raleigh","Bismarck","Columbus","Oklahoma_City","Salem","Harrisburg","Providence","Columbia","Pierre","Nashville","Austin","Salt_Lake_City","Montpelier","Richmond","Olympia","Charleston","Madison","Cheyenne")
visited <- c("Honolulu", "Nashville", "Boston", "Amherst", "San Diego")
visited %in% cities
visited[visited %in% cities]
```


#### Functions
The above pieces of code can be run independently, but to reuse them (or pieces of code in general) efficiently, they should be put into functions. A `function()` is an R object of class _function_ that shares some characteristics with other objects: it can be passed as an argument to a _function_ and it can be within another _function_. To create a function, we specify what should be input, perform some computation on the input, and return an output. The structure of a function in R is the following:
```{r, eval=FALSE}
f <- function( arguments ){
  #Code to do something
}
```
The arguments that get passed to `[arguments]` are called _formal_ arguments. They can be passed based on their position or by name. To avoid confusion, stick to the given order as much as possible and use names, especially for arguments that have default values. Here are some simple example functions to go over general structure of functions. We will learn more about functions throughout the course.
```{r, eval=FALSE}
add <- function(x, y){ #input is x and y
  z <- x+y #computation perfomed
  z #output - the default is that last thing in function gets returned
}
add(2, 3)

is.negative <- function(x){
  if(x < 0) TRUE else FALSE
}
is.negative(5)
```
Here are some things to notice about arguments using the `mean()` function. Call `?mean` and notice that it has one listed formal argument, `x` and two named arguments with default values. The `...` portion refers to arguments that are inherited from other methods or arguments that can be passed on to another method. We will not worry about that now.
```{r, eval=TRUE}
x <- c(1:20, NA, 21:50)
mean(x) #Uses default values
mean(x, na.rm=TRUE)
mean(x, na.rm=TRUE, trim=0.1) #Note that order change of na.rm and trim doesn't affect outcome since names used
```
You can create a function using formal and named arguments as follows:
```{r, eval=TRUE}
s.calc <- function(x, y, a=1, b=2){ #input is x and y
  z <- a*x + b*y #computation perfomed
  z #output - the default is that last thing in function gets returned
}
s.calc(1, 1)
s.calc(1, 2, a=0)
s.calc(1, 2, b=0)
s.calc(1, 2, a=0, b=0)
```


### Scoping
Notice that `a` and `b` in the previous chunk have known values only within the function by typing `a` and `b` in the Console. If we set `a <- 10` before the function and re-run the code, the output is the same notice what changes:
```{r, eval=TRUE}
a <- 10
s.calc(1, 2, a=0, b=0)
a
```
Now try these to emphasize the difference between the global and function environments:
```{r, eval=TRUE}
a <- 10
s.calc <- function(x, y){ #input is x and y
  z <- a*x + y 
  z 
}
s.calc(1, 1)
a
s.calc <- function(x, y){ 
  a = 0
  z <- a*x + y 
  z 
}
s.calc(1, 1)
a
```

R uses _static scoping_ which means all objects used by a function must be visible in the environment in which the function is defined. Let's see what happens with the example mentioned during lecture.
```{r, eval=TRUE}
a <- 0
b <- 1
f <- function(x) {a*x + b}
f(2)
g <- function(x){
    a <- 1
    b <- 2
    f(x)
}
g(2)
c(a, b)
g <- function(x){
    a <- 1
    b <- 2
    print(c(a,b))
    print(f(x)) #Will compute f(x) in the environment where f(x) is defined
    a*x + b #The same function as f(x) except computed within this environment
}
g(2)
c(a, b)
```


R allows functions to be defined within functions, which means that a function can return a function and there can be several levels of environments. 
```{r, eval=TRUE}
s.calc.fun <- function(x){ #input is x and y
  out.fun <- function(y){
    2*x + y 
  } 
  out.fun #output is now a function
}
f1 <- s.calc.fun(1)
f2 <- s.calc.fun(2)
f1
f2
f1(1)
f2(1)
```
To see what is within a function's environment, you can use:
```{r, eval=TRUE}
ls(environment(f1))
get("x", environment(f1))
get("out.fun", environment(f1))
environment(s.calc.fun)
ls(environment(s.calc.fun)) #Same as ls()
```
Recall that `<-` and `=` can behave differently. When defining variables in function argument lists, use `=` to avoid setting global variables. For example, compare:
```{r, eval=TRUE}
x <- 1:4
y <- c("a", "b", "c", "d")
single <- list(x=1, y="a")
single
x
y
single <- list(x<-1, y<-"a")
single
x
y
```


### Symbol Binding
The order in which R looks for symbols can be seen with `search()`. Note that `.GlobalEnv` refers to the global environment, which is the set objects that the user has called in for a given session. Packages are loaded in a specific order: the most recent packages are loaded such that their objects replace pre-existing ones. 
```{r, eval=TRUE}
search()
```
To avoid potential errors when there are functions with the same name in different packages, you can explicitly state the package from which a function should be called. This can also be good practice for identifying package dependency-related errors and writing clean code. For example, type ``purrr:map`` rather than ``map`` to indicate you are using the map function from the purrr package.




